---
title: "Mathematical Annotation in R"
output: html_notebook
---

In this notebook, we demonstrate how to include mathematical symbols and formulas in plots created with R.

Base R provides a LaTeX-like language for defining mathematical expressions. They are documented on the `plotmath` help page, which can be accessed by typing in the R console `?plotmath`. This language can be used in either base R graphics or ggplot2.

plotmath expressions can be used in titles, subtitles, axis labels, and annotations within the plot.

To create a mathematical expression using the plotmath language, we need to place the expression inside the `expression()` function. 

## Greek symbols

Here's a simple plot with no mathematical notation. It's a standard Normal distribution.

```{r}
curve(dnorm(x), from = -3, to = 3)
```

Now we add an expression to the title of the plot: "mu == 0". Two equal signs is rendered as a single equals sign. The keyword mu is rendered as a lower case Greek mu. 

```{r}
curve(dnorm(x), from = -3, to = 3,
      main = expression(mu == 0))
```

All letters of the Greek alphabet are available to use:

- alpha -- omega (lower case)
- Alpha -- Omega (upper case)

Equalities are defined the same as they are in R code:

- x == y
- x != y (not equal)
- x < y	
- x <= y	
- x > y	
- x >= y	

Other common expressions

- x^2	  
- sqrt(x)
- sqrt(x, y)	(yth root of x)
- x*y	(juxtapose x and y)
- x %+-% y	(x plus or minus y)
- x %*% y	(x times y)
- x[i]	(x subscript i)


## Exercise 1

The following is an exponential distribution with lambda = 1. Add a title to the plot that says lambda = 1, where lambda is rendered as a lower case Greek symbol. Use the `main` argument to add the title.

```{r}
curve(dexp(x), from = 0, to = 6, main = expression(lambda == 1))
```

## Combining symbols and text

To mix text and math symbols, we use the `paste()` function within `expression()`. For example, let's include "and sigma == 1" in the plot title for our standard normal curve.

```{r}
curve(dnorm(x), from = -3, to = 3,
      main = expression(paste(mu == 0, " and ", sigma == 1)))
```


## Exercise 2

Add a title to the following plot that says lambda = 1 and x >= 0. Use the `main` argument to add the title.

```{r}
curve(dexp(x), from = 0, to = 6, 
      main = expression(paste(lambda == 1, " and ", x >= 0)))

```

## Building plotmath expressions

The previous two examples were simple. However math expressions can often be complicated and take time to create. It can help to have a "sandbox" to gradually build and test your plotmath expressions. Here's one way using `plot()` and `text()`.

`plot(1,1, type = "n")` creates an empty plot. 
`text(1,1, expression())` will plot your expression in the middle of the plot.

Once you're done with your expression, you can copy and paste it into the R code where needed. Or save as an object.

```{r}
plot(1,1, type = "n")
text(1,1, expression(y == beta[0] + beta[1]*x + epsilon))
```

Let's say we're happy with that. We can save it.

```{r}
e <- expression(y == beta[0] + beta[1]*x + epsilon)
```

And then use it in a plot. For example.

```{r}
m <- lm(dist ~ speed, cars)
plot(dist ~ speed, cars, main = e)
abline(m)
text(7, 80, e)
```

## Exercise 3

```{r}
plot(1, 1, type = "n")
text(1, 1, expression(A == pi * r^2))
```



## plotmath with ggplot2

ggplot2 also allows plotmath expressions, but how use them depends on where you want them. 

- In `annotate()`, `geom_text()` and `geom_label()`, they have to be strings, not expressions, and include the argument `parse = TRUE`.
- In `labs()` and `ggtitle()` use plotmath expressions just as you would in base R.

```{r message=FALSE}
library(ggplot2)
e <- expression(y == beta[0] + beta[1]*x + epsilon)
e2 <- "y == beta[0] + beta[1]*x + epsilon"
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", 7, 80, label = e2, parse = TRUE) +
  labs(title = e) 
```

To change the size of your expression:

- In `annotate()`, `geom_text()` and `geom_label()`, use the `size` argument to specify point size in mm. 
- for expressions in `labs()` and `ggtitle()`, you need to use the `theme()` function and modify the appropriate argument with `element_text(size = x)`, where x is the point size as specified in programs like MS Word. 

For example:

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", 7, 80, label = e2, parse = TRUE, size = 10) + 
  labs(title = e) +
  theme(title = element_text(size = 12))
```

```{r}
ggplot(iris) +
  aes(x = Petal.Width, y = Petal.Length, color = Species) +
  geom_point() +
  scale_color_discrete(expression(mu), 
                       labels = list(expression(mu*1), 
                                     expression(mu*2), 
                                     expression(mu*3)))
```

```{r}
n <- 200
g <- gl(n = 3, k = n)
set.seed(1)
y <- c(rnorm(n, sd = 5),
       rnorm(n, sd = 10),
       rnorm(n, sd = 20))
d <- data.frame(y, g)
```

```{r}
ggplot(d) +
  aes(x = y, fill = g) +
  geom_density(alpha = 1/3)
```

```{r}
ggplot(d) +
  aes(x = y, fill = g) +
  geom_density(alpha = 1/3) +
  scale_fill_discrete(expression(sigma), labels = 
                        list(expression(sigma == 5),
                             expression(sigma == 10),
                             expression(sigma == 20)))
```

```{r}
library(latex2exp)

TeX(r"($y = \beta_0 + \beta_1 x + \epsilon$)", output = "character")

```

$$y = \beta_0 + \beta_1 x + \epsilon$$

```{r}
ggplot() +
  geom_function(aes(color = "1"), fun = dgamma, args = list(shape = 1, rate = 1/4)) +
  geom_function(aes(color = "2"), fun = dgamma, args = list(shape = 2, rate = 1/4)) +
  geom_function(aes(color = "3"), fun = dgamma, args = list(shape = 3, rate = 1/4)) +
  xlim(0, 30) +
  scale_color_discrete(expression(alpha))
```

```{r}
s <- matrix(c(1,0.5,0.5,1), ncol = 2)
set.seed(2)
d2 <- MASS::mvrnorm(n = 75, mu = c(0,0), Sigma = s) |> as.data.frame()
r <- cor(d2)[1,2]
ggplot(d2) +
  aes(V1, V2) +
  geom_point() +
  labs(title = bquote(rho == .(r)))
```


```{r}
e <- rexp(100, rate = 7)
r <- 1/mean(e)
hist(e, main = bquote(hat(lambda) == .(r)))
```

```{r}
m <- lm(dist ~ speed, data = cars)
summary(m)

```


```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = expression(beta[1] == 3.93 %+-% 0.84)) +
  annotate("text", 7, 80, label = "R^2 == 0.64", parse = TRUE)
```

